COVID19 - Forecast and predictions using a BYM model in Italy

### 2020-03-24 BYM_model/INLA_def.Rmd
Add home page, navigation and theme

---
title: "COVID19 - Forecast and predictions using a BYM model - Veneto Region"
author: "Paolo Girardi"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: rmarkdown::github_document
always_allow_html: yes
---
```{r echo=FALSE}
#Load libraries
library(spdep)
library(maptools)
library(INLA)
library(rgdal)
library(foreign)
library(dygraphs)
library(xts)
```
################################ IMPORT DATASETS
# ISTAT Province dataset
```{r}
db_istat<-read.csv("Elenco-comuni-italiani2.csv",sep=";",header=T,dec=",")
pop_provincia<-data.frame(codice_provincia=as.numeric(names(table(db_istat[,3]))),pop=tapply(db_istat[,20],db_istat[,3],sum))
###import international dataset
dat_int<-read.csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_19-covid-Confirmed.csv",header=T)
hubei<-as.numeric(dat_int[dat_int$Province.State=="Hubei",5:dim(dat_int)[2]])

###import updated italian dataset  for each province
dat_csv<-read.csv("https://raw.githubusercontent.com/pcm-dpc/COVID-19/master/dati-province/dpc-covid19-ita-province.csv",header=T)
dat_csv<-merge(dat_csv,pop_provincia,by=c("codice_provincia"),all.y=TRUE)
dat_csv<-dat_csv[dat_csv$codice_provincia<112,]
#select a Region, in this example "Lombardia" Code region 3
Region<-3
dat_csv<-dat_csv[dat_csv$codice_regione==Region,]
dat_csv_n$denominazione_provincia<-droplevels(dat_csv_n$denominazione_provincia)
#### number of province
nprov<-length(table(dat_csv$denominazione_provincia)[table(dat_csv$denominazione_provincia)>0])

#order the obsercation for date and space, useful for INLA later....
dat_csv<-dat_csv[order(dat_csv$data),]
dat_csv<-dat_csv[order(dat_csv$codice_provincia),]
###space ID
dat_csv$ID<-as.factor(dat_csv$codice_provincia)
levels(dat_csv$ID)<-1:length(levels(dat_csv$ID))
dat_csv$ID<-as.numeric(dat_csv$ID)
#calculate new_cases
dat_csv$cumulative_cases<-dat_csv$totale_casi
dat_csv$totale_casi[dat_csv$data!="2020-02-24 18:00:00"]<-unlist(tapply(dat_csv$totale_casi,dat_csv$ID,diff))
dat_csv$totale_casi[dat_csv$totale_casi<0]<-0

#import dataset shapefile Italian province
nc.province <- readShapePoly(paste(wd,"/ProvCM01012019_g/ProvCM01012019_g_WGS84.shp",sep=""))
nc.province<-nc.province[nc.province$COD_REG==Region,]
nc.province<-nc.province[order(nc.province$COD_PROV),]
nc.province$ID<-as.factor(nc.province$COD_PROV)
levels(nc.province$ID)<-1:length(levels(nc.province$ID))
nc.province$ID<-as.numeric(nc.province$ID)
#Create adjacency matrix
nc.nb <- poly2nb(nc.province,snap=0.01)
nb2INLA(paste(wd,"nc.adj",sep=""), nc.nb)
file.adj <- paste(wd,"nc.adj",sep="")
################################
```

################################
#prior distribution for iid and besag components
```{r}
prior.iid = c(1,0.01)
prior.besag = c(1,0.001)
initial.iid = 4
initial.besag = 3
```
####COVID19 MAPS BYM with INLA
##########################indexing, for 
#### number of days
```{r}
dat_csv$t<-as.numeric(dat_csv$data)
days<-max(dat_csv$t)
dat_csv$t3<-dat_csv$t2<-dat_csv$t
#### province
dat_csv$ID2<-dat_csv$ID
##########################
formula.bym= totale_casi ~ 1+f(ID, model = "bym", graph = file.adj, param = c(prior.iid, prior.besag), initial = c(initial.iid, initial.besag))
fit_1<-inla(formula.bym, family="poisson", data=dat_csv, E=pop,control.compute = list(dic=T))
summary(fit_1)
##incidence rate ratio for each province
IRR_mean <- exp(fit_1$summary.random$ID$mean[1:nprov])
nc.province$IRR_mean<-IRR_mean
spplot(nc.province, c( "IRR_mean"))
```
#Pure temporal model (RW2)
RW2 temporal model
```{r}
Date<-seq(as.Date("2020-02-24"),as.Date("2020-02-24")+days-1,1)
formula_t = totale_casi ~ 1+f(t,model="rw2", constr = FALSE)+f(t2,model="iid")
fit_2<-inla(formula_t, family="poisson", data=dat_csv, E=pop,control.compute = list(dic=T))
summary(fit_2)
plot(Date,fit_2$summary.random$t$mean,ylab="Temp Coef",xlab="Date")
#estimating a linear model without the first obs
lm_fit<-lm(fit_2$summary.random$t$mean~I(1:days),subset=-1)
coef(lm_fit)[2]
plot(1:days,fit_2$summary.random$t$mean,ylab="Temp Coef",xlab="Date")
abline(lm_fit)
##Incidence rate ratios, the rate of increase od cases respect to time 1
plot(Date,exp(fit_2$summary.random$t$mean),ylab="Indicence of cases",xlab="Date")
```

#interaction model type 4 (complete time trend interaction to estimate specific curves for each province)
```{r}
####specification
formula.intIV<- totale_casi ~f(ID,model="bym",graph=file.adj) +
  f(t,model="rw2",constr = FALSE) +
  f(t2,model="iid") +
  f(ID2,model="besag", graph=file.adj,
    group=t3,
    control.group=list(model="rw2"))
#estimation
fit_st4<-inla(formula.intIV, family="poisson", data=dat_csv, E=pop,control.compute = list(dic=T))
summary(fit_st4)
```
#### components
```{r}
t_1<-fit_st4$summary.random$t$mean #overall trend
t_2<-fit_st4$summary.random$t2$mean #iid trend
s_1<-fit_st4$summary.random$ID$mean[1:nprov] # spatial CAR component
s_2<-fit_st4$summary.random$ID$mean[(nprov+1):(nprov*2)] # spatial idd
st<-fit_st4$summary.random$ID2$mean # spatio-temporal model with car and ar1 specification ```
## Mean IRR for each province
```{r}
nc.province$IRR_st<-exp(s_1+apply(t(matrix(st,nrow=nprov)),2,mean))
spplot(nc.province, c( "IRR_st"),main="IRR")
```
## Trends for province
```{r}
trends<-t_1+t(matrix(st,nrow=nprov))
# trends+ specific
matplot(trends)
#specific for each province
matplot(t(matrix(st,nrow=nprov)))
```

##################predictions 3 days forward of new CODID 19 cases
```{r}
Forecast=3
dat_csv2<-dat_csv 
dat_csv2$t<-rep(1:days+days,nprov)
dat_csv2$totale_casi<-NA
dat_csv2<-dat_csv2[dat_csv2$t<=(days+Forecast),]
dat_csv_n<-rbind(dat_csv,dat_csv2)
dat_csv_n$t3<-dat_csv_n$t2<-dat_csv_n$t
#introducting hubei starting on 22/01, quarantine started on 25/01
# applying a smoothing , data has errors
hubei_new_cases<-filter(diff(hubei), rep(1, 5))

plot(hubei/58.5,type="l",ylab="COVID19 cases for milion of inhabitant",ylim=c(0,1530),xlab="Days since over 100 cases")
lines(c(rep(NA,2),cumsum(tapply(dat_csv_n$totale_casi,dat_csv_n$t,sum))/10),col=2)
abline(v=2,col=1,lty=2)
abline(v=4,col=2,lty=2)
abline(v=10,col=3,lty=3)
legend("bottomright",c("Hubei","Lombardy"),lty=1,col=1:2)
text(10,1500,"Hubei quarantine")
text(12,1350,"Lodi home lock",col=2)
text(20,1200,"Lombardy home lock",col=3)
#province of lodi restrictive measure started on 24/02 lag 28 days, at the beginning of the time series
#other provinces on 11/03 lagged started their restrictions
hubei_lodi<-hubei_new_cases[1:(days+Forecast)]
hubei_others<-c(rep(NA,8),hubei_new_cases[1:(days+Forecast-8)])
dat_csv_n<-dat_csv_n[order(dat_csv_n$codice_provincia,dat_csv_n$t),]
dat_csv_n$hubei<-rep(hubei_others,12)
dat_csv_n$hubei[dat_csv_n$denominazione_provincia=="Lodi"]<-hubei_lodi
#overdispersion
dat_csv_n$over<-1:dim(dat_csv_n)[1]
```
We model the number of cases.

In the rw2 I set constr = FALSE is set to FALSE and that, for this reason, the intercept is not included in the linear predictor. 
```{r}
## include overdispersion parameter
formula.intIVn<- totale_casi ~ log(hubei+1)+log(pop)+
  f(t,model="rw2", constr = FALSE) +
  f(t2,model="iid") +f(over,model="iid") +
f(ID,model="bym",graph=file.adj) +
  f(ID2,model="besag", graph=file.adj,
    group=t3,
    control.group=list(model="rw2"))+ f(over,model="iid")

fit_st4n<-inla(formula.intIVn, family="poisson", data=dat_csv_n,control.compute = list(dic=T),control.predictor = list(link = 1))
summary(fit_st4n)
```

```{r}
#### components
t_1<-fit_st4n$summary.random$t$mean #overall trend
t_2<-fit_st4n$summary.random$t2$mean #iid trend
s_1<-fit_st4n$summary.random$ID$mean[1:nprov] # spatial CAR component
s_2<-fit_st4n$summary.random$ID$mean[(nprov+1):(nprov*2)] # spatial idd
st<-fit_st4n$summary.random$ID2$mean # spatio-temporal coefficients 
## Trends for province
Date_n<-seq(as.Date("2020-02-24"),as.Date("2020-02-24")+days+Forecast-1,1)
trends<-t_1+t(matrix(st,nrow=nprov))
plot(Date_n,t_1,ylab="RW2 Time coefficient",xlab="Date")
matplot(t(matrix(st,nrow=nprov)),ylab="RW2 Time coefficient for Province",xlab="Date")
#extract number of cases for lodi
est_lodi<-fit_st4n$summary.fitted.values[dat_csv_n$denominazione_provincia=="Lodi",]

>>>>

---
title: "COVID19 - Forecast and predictions using a BYM model in Italy"
author: "Paolo Girardi"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: rmarkdown::github_document
always_allow_html: yes
---
```{r echo=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir= normalizePath('..'))
knitr::opts_chunk$set(error = FALSE)
```
<a rel="license" href="http://creativecommons.org/licenses/by-nc/4.0/"><img alt="Creative Commons License" style="border-width:0" src="https://i.creativecommons.org/l/by-nc/4.0/88x31.png" /></a><br />This work is licensed under a <a rel="license" href="http://creativecommons.org/licenses/by-nc/4.0/">Creative Commons Attribution-NonCommercial 4.0 International License</a>.

# Disclaimer

* We want to investigate the evolution of the coronavirus pandemic in  Italy

* Our point of view is that of surveillance with the goal of detecting important changes in the underlying (random) process as soon as possible after it has occured.

* We use data provided by  Italian Civil Protection Department and the analysis was restricted to the Lombardy Region

* This document is in a draft mode, and it is continuously updated.

* The layout of the draft must definitely be improved.

*NB: set the file output format to

#output:html_document:  
  df_print: paged  
pdf_document:   
  toc: yes

which  performs the same analysis enabling Javascript Pictures.

## The COVID dataset

The present analysis started from the dataset on COVID19 updated in
<https://github.com/pcm-dpc/COVID-19>, database provided by the Italian
Civil Protection.

# Software

Install  packages `dygraphs`,  `xts` and `EpiDynamics` if not available

```{r echo=TRUE}
checkpackage <- function(package) {
  if (!package %in% installed.packages()) install.packages(package)
}
checkpackage("dygraphs")
checkpackage("xts")
checkpackage("ggplot2")
checkpackage("spdep")
checkpackage("maptools")
checkpackage("INLA")
checkpackage("rgdal")
```
and load them.
```{r, warning=FALSE}
#Load libraries
library("dygraphs")
library("xts")
library("ggplot2")
library("spdep")
library("maptools")
library("INLA")
library("rgdal")
```
# Datasets loading
```{r}
db_istat<-read.csv("https://raw.githubusercontent.com/Paolin83/COVID19_SEIR_model/master/BYM_model/Elenco-comuni-italiani2.csv",sep=";",header=T,dec=",")
pop_provincia<-data.frame(codice_provincia=as.numeric(names(table(db_istat[,3]))),pop=tapply(db_istat[,20],db_istat[,3],sum))
###import international dataset
dat_int<-read.csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_19-covid-Confirmed.csv",header=T)
hubei<-as.numeric(dat_int[dat_int$Province.State=="Hubei",5:dim(dat_int)[2]])

###import updated italian dataset  for each province
dat_csv<-read.csv("https://raw.githubusercontent.com/pcm-dpc/COVID-19/master/dati-province/dpc-covid19-ita-province.csv",header=T)
dat_csv<-merge(dat_csv,pop_provincia,by=c("codice_provincia"),all.y=TRUE)
dat_csv<-dat_csv[dat_csv$codice_provincia<112,]
#select a Region, in this example "Lombardia" Code region 3
Region<-c(3)
dat_csv<-dat_csv[dat_csv$codice_regione %in%Region,]
dat_csv$denominazione_provincia<-droplevels(dat_csv$denominazione_provincia)
#### number of province
nprov<-length(table(dat_csv$denominazione_provincia)[table(dat_csv$denominazione_provincia)>0])

#order the obsercation for date and space, useful for INLA later....
dat_csv<-dat_csv[order(dat_csv$data),]
dat_csv<-dat_csv[order(dat_csv$codice_provincia),]
###space ID
dat_csv$ID<-as.factor(dat_csv$codice_provincia)
levels(dat_csv$ID)<-1:length(levels(dat_csv$ID))
dat_csv$ID<-as.numeric(dat_csv$ID)
#calculate new_cases Y ~ Poi(mu)
dat_csv$cumulative_cases<-dat_csv$totale_casi
dat_csv$totale_casi[dat_csv$data!="2020-02-24 18:00:00"]<-unlist(tapply(dat_csv$totale_casi,dat_csv$ID,diff))
dat_csv$totale_casi[dat_csv$totale_casi<0]<-0
#import dataset shapefile Italian province and adjancies matrix
#### IMPORTANT, SET THE Work Directory of the shapefile 
setwd("/Users/Paolo/My_Website/public/COVID19_SEIR_model")
nc.province <- readShapePoly("BYM_model/ProvCM01012019_g/ProvCM01012019_g_WGS84.shp")
nc.province<-nc.province[nc.province$COD_REG %in% Region,]
nc.province<-nc.province[order(nc.province$COD_PROV),]
nc.province$ID<-as.factor(nc.province$COD_PROV)
levels(nc.province$ID)<-1:length(levels(nc.province$ID))
nc.province$ID<-as.numeric(nc.province$ID)
#Create adjacency matrix
nc.nb <- poly2nb(nc.province,snap=0.01)
nb2INLA("nc.adj", nc.nb)
file.adj <- "nc.adj"
################################
```

# Spatial analysis with R-INLA and BYM   
We modelled COVID 19 cases by means of a BYM (Besag, York and Molli\'e) model using an Integrated Nested Laplace Approximation (INLA).   
\textbf{Starting phase}

For the $i$-th Nuts-3 Region (Italian province), the observed number of COVID19, \textbf{$y_{i}$}, was modelled as follows:  
$$
y_{i} \sim Poisson (\lambda_{i})\\
$$
with i=1,..., `r nprov`. We modelled the incidence of COVID-19 by means of a BYM model including the population size as offset variable as:  
$$
	log(\frac{\lambda_{i}}{N_{i}})=\alpha+\mu_i+\nu_i,\\
$$
where $\alpha$ is the intercept, $\mu_i$ and $\nu_i$ are two area specific effects with normal distribution modelled using an intrinsic conditional autoregressive structure (iCAR) and $N_i$ is the population size of each NUTS-3 Region.

The parameter were estimated by INLA (Integrated Nested Laplace Approximation) and R software.

```{r}
nomi_prov<-levels(dat_csv$denominazione_provincia)
barplot(tapply(dat_csv$totale_casi,dat_csv$denominazione_provincia,sum),names =1:12,args.legend = list(x = "topright"),legend.text =paste(1:12,nomi_prov),ylab="Number fo cases")
```
The number of observed cases $y_i$ is very different among  NUTS-3 regions.

```{r}
pop<-tapply(dat_csv$pop,dat_csv$denominazione_provincia,mean)
barplot(tapply(dat_csv$totale_casi,dat_csv$denominazione_provincia,sum)/pop*1000,names =1:12,args.legend = list(x = "topright"),legend.text =paste(1:12,nomi_prov),ylab="Number of cases x 1.000 inhabitants")
```
A correct comparison is the made by incidence cases $y_i/N_i$, here riported for x1000 inhabitants. 
#Spatial model  
We estimate a BYM model fixing prior distributions for iid and besag components as follows:  
```{r}
prior.iid = c(1,0.01)
prior.besag = c(1,0.001)
initial.iid = 4
initial.besag = 3
```
We create new variables that are required by INLA procedure.  
```{r}
##########################indexing, for 
#### number of days
dat_csv$t<-as.numeric(dat_csv$data)
days<-max(dat_csv$t)
dat_csv$t3<-dat_csv$t2<-dat_csv$t
#### province
dat_csv$ID2<-dat_csv$ID
#### province
dat_csv$IDt<-1:dim(dat_csv)[1]
########################## formula BYM
formula.bym= totale_casi ~ 1+f(ID, model = "bym", graph = file.adj, param = c(prior.iid, prior.besag), initial = c(initial.iid, initial.besag))
fit_1<-inla(formula.bym, family="poisson", data=dat_csv, E=pop,control.compute = list(dic=T))
summary(fit_1)
##incidence rate ratio for each province
IRR_mean <- exp(fit_1$summary.random$ID$mean[1:nprov])
nc.province$IRR_mean<-IRR_mean
nc.province$DEN_PROV<-droplevels(nc.province$DEN_PROV)

spplot(nc.province, c( "IRR_mean"),main="Incidence Rate Ratios", sp.layout=
         list("sp.text", coordinates(nc.province), nc.province$DEN_PROV, cex=0.7, col="gray"))

```
The image reported the IRR=exp($\mu_i$), the increase of the Incidence of Covid-19 in each NUTS-3 Region respect to the overall mean.

#Temporal model   
We consider the temporal aspect estimating a Bayesian RW2 model to model the temporal trend of COVID-19 in the reported temporal window. The observed number of COVID-19 a the time $t$, \textbf{$y_{t}$}, is modelled as follows  
$$
y_{t} \sim Poisson (\lambda_{t})
$$
with t=1,...,`r days`.
\end{center}
where the quantity $\frac{\lambda_{t}}$, is modelled by a Random Walk of order 2 variable
$$
log({\lambda_{it}})=\alpha+\gamma_t+\phi_t, 
$$
where $\alpha$ is the intercept and $\gamma_t \sim RW2$  are the coefficients related to the random walk process while $\phi_t \sim N(0, \tau_t^{-1} )$ are the temporal specific random errors.
```{r}
Date<-seq(as.Date("2020-02-24"),as.Date("2020-02-24")+days-1,1)
plot(Date,tapply(dat_csv$totale_casi,dat_csv$t,sum),xlab="Date",ylab="New cases",type="l")
points(Date,tapply(dat_csv$totale_casi,dat_csv$t,sum))
formula_t = totale_casi ~ 1+f(t,model="rw2", constr = FALSE)+f(t2,model="iid")
fit_2<-inla(formula_t, family="poisson", data=dat_csv,control.compute = list(dic=T))
summary(fit_2)

##Incidence rate ratios, increase od cases respect to time 1
plot(Date,exp(fit_2$summary.random$t$mean),ylab="temporal IRR",xlab="Date")

label<-Date

mean  <- exp(fit_2$summary.random$t$mean[1:days]+fit_2$summary.fixed$mean)*nprov
lower <- exp(fit_2$summary.random$t$mean[1:days]-1.96*fit_2$summary.random$t$sd[1:days]+fit_2$summary.fixed$mean)*nprov
upper <- exp(fit_2$summary.random$t$mean[1:days]+1.96*fit_2$summary.random$t$sd[1:days]+fit_2$summary.fixed$mean)*nprov

df <- data.frame(label, mean, lower, upper)

fp <- ggplot(data=df, aes(x=label, y=mean, ymin=lower, ymax=upper)) +
  geom_pointrange() +
  geom_hline(yintercept=1, lty=2) +  # add a dotted line at x=1 after flip
  xlab("Date") + ylab("Estimated new cases") +
  theme_bw() 
print(fp)

```
We now consider a BYM model with spatial and temporal components and their complete interactions (Type IV). The observed number of COVID-19, \textbf{$y_{it}$}, was modelled as:
$$
y_{it} \sim Poisson (\lambda_{it})\\
$$
with i=1,..,`r nprov`, and t=1,...,`r days`. We modelled the ratio $\frac{\lambda_{it}}{N_{i}}$, the Incidence of COVID-19, with a BYM model including temporal covariates as follows  
$$
log(\frac{\lambda_{it}}{N_{i}})=\alpha+\mu_i+\nu_i+\gamma_t+\phi_t+\delta_{it},\\
$$

where $\alpha$ is the intercept and:  
*space: $\mu_i$ and $\nu_i$ are two area specific effects with normal distribution modelled using an intrinsic conditional autoregressive structure (iCAR);

*time: $\gamma_t$ are the coefficients related to the time process, while $\phi_t \sim N(0, \tau_t^{-1} )$ are the temporal specific random errors. Time process can be autoregressive or a random walk process (order 1 or 2);

*space-time: $\delta_{it}$ are the coefficients related space-time interactions.

There are many type of spatio-temporal interactions (please see Spatial and Spatio-temporal Bayesian Models with R - INLA, Blangiardo and Cameletti).

We explored four spatio-temporal interactions:
-Type I:  coefficients $\delta_{it}$ do not have a spatial or/and temporal structure ($\delta_{it}\sim N(0, 1/\tau_{\delta})$).
-TypeII: coefficients $\delta_{it}$ have a structured temporal main effect and the unstructured spatial effect.  
-TypeIII: coefficients $\delta_{it}$ have a spatial structure (as Besag), but are temporally unstructured.
-TypeVI: coefficients $\delta_{it}$ have both a spatial  and temporal structure. It is the most complex type of space-time interaction.

```{r}
####specification
formula.intI<-  totale_casi ~ 1+ f(ID,model="bym",graph=file.adj) +
f(t,model="rw2") +f(t2,model="iid") +
f(IDt,model="iid")

formula.intII<-  totale_casi ~ 1+f(ID,model="bym",graph=file.adj) +
f(t,model="rw2") + f(t2,model="iid") +
f(ID2,model="iid", group=t3,control.group=list(model="rw2"))

formula.intIII<-   totale_casi ~ 1+f(ID,model="bym",graph=file.adj) +
f(t,model="rw2") + f(t2,model="iid") +
f(t3,model="iid", group=ID2, control.group=list(model="besag", graph=file.adj))

formula.intIV<- totale_casi ~ 1+f(ID,model="bym",graph=file.adj) +
f(t,model="rw2") + f(t2,model="iid") +
f(ID2,model="besag", graph=file.adj, group=t3,control.group=list(model="rw2"))

#estimation
fit_st1<-inla(formula.intI, family="poisson", data=dat_csv, E=pop,control.compute = list(dic=T))
fit_st2<-inla(formula.intII, family="poisson", data=dat_csv, E=pop,control.compute = list(dic=T))
fit_st3<-inla(formula.intIII, family="poisson", data=dat_csv, E=pop,control.compute = list(dic=T))
fit_st4<-inla(formula.intIV, family="poisson", data=dat_csv, E=pop,control.compute = list(dic=T))

# best DIC
fit_st1$dic$dic
fit_st2$dic$dic
fit_st3$dic$dic
fit_st4$dic$dic
### the first model is the best, but we choice the third... because we want to force the spatial structure in the delta_ij.
```

  The time and space component mean estimates can be extracted here

```{r}
#overall trend
t_1<-fit_st3$summary.random$t$mean 
#iid trend
t_2<-fit_st3$summary.random$t2$mean 
 # spatial iCAR component
s_1<-fit_st3$summary.random$ID$mean[1:nprov]
# spatial idd
s_2<-fit_st3$summary.random$ID$mean[(nprov+1):(nprov*2)] 
#spatio-temporal model with car and ar1 specification
st<-fit_st3$summary.random$t3$mean

```

When can represent the IRR(=exp($\mu_i+\frac{1}{T} \sum_{i=t}^{T} \delta_{it}$)) for each NUTS-3 regions.

```{r}
nc.province$IRR_st<-exp(s_1+apply(t(matrix(st,nrow=days)),1,mean))
l1 = list("sp.text", coordinates(nc.province), nc.province$DEN_PROV,col="grey", cex=0.7,font=2)
spplot(nc.province, c( "IRR_st"),main="Incidence ", sp.layout=list(l1))
```

Trend for each province.

```{r}
##overall trend
plot(Date,exp(t_1),ylab="IRR",main="Overall time trend")

## time trend for each province
db_plot<-data.frame(t_1+matrix(st,nrow=days))
names(db_plot)<-nc.province$DEN_PROV
names(db_plot)[4]<-"Milano"
province_trend<-xts(db_plot, order.by = Date, frequency = 7)
p <- dygraph(province_trend,main=paste("Lombardy Region",sep =""),xlab="Day",height=400,width=800,ylab= "Province trends") 
p

#space time variation for each province from overall trend
db_plot<-data.frame(matrix(st,nrow=days))
names(db_plot)<-nc.province$DEN_PROV
names(db_plot)[4]<-"Milano"
province_trend<-xts(db_plot, order.by = Date, frequency = 7)
p <- dygraph(province_trend,main=paste("Lombardy Region",sep =""),xlab="Day",height=400,width=800,ylab= "Delta coefficient") 
p
```

```{r}
q4<-round(quantile(1:days,0:4/4),0)

#Incidence divided by four

Incidence_1q<-tapply(fit_st3$summary.fitted.values$mean[dat_csv$t %in%q4[1]:(q4[2]-1)], dat_csv$ID[dat_csv$t %in%q4[1]:(q4[2]-1)],sum)
Incidence_2q<-tapply(fit_st3$summary.fitted.values$mean[dat_csv$t %in%q4[2]:(q4[3]-1)], dat_csv$ID[dat_csv$t %in%q4[2]:(q4[3]-1)],sum)
Incidence_3q<-tapply(fit_st3$summary.fitted.values$mean[dat_csv$t %in%q4[3]:(q4[4]-1)], dat_csv$ID[dat_csv$t %in%q4[3]:(q4[4]-1)],sum)
Incidence_4q<-tapply(fit_st3$summary.fitted.values$mean[dat_csv$t %in%q4[4]:(q4[5])], dat_csv$ID[dat_csv$t %in%q4[4]:(q4[5])],sum)
nc.province$Incidence1<-Incidence_1q
nc.province$Incidence2<-Incidence_2q
nc.province$Incidence3<-Incidence_3q
nc.province$Incidence4<-Incidence_4q
cols = bpy.colors(100)
col_lim<-range(c(Incidence_1q,Incidence_2q,Incidence_3q,Incidence_4q))
c1i<-c(round((range(Incidence_1q)[1]/col_lim[2])*100))+1
c1f<-c(round((range(Incidence_1q)[2]/col_lim[2])*100))
c2i<-c(round((range(Incidence_2q)[1]/col_lim[2])*100))+1
c2f<-c(round((range(Incidence_2q)[2]/col_lim[2])*100))
c3i<-c(round((range(Incidence_3q)[1]/col_lim[2])*100))+1
c3f<-c(round((range(Incidence_3q)[2]/col_lim[2])*100))
c4i<-c(round((range(Incidence_4q)[1]/col_lim[2])*100))+1
c4f<-c(round((range(Incidence_4q)[2]/col_lim[2])*100))

plot<-list()
i<-1
p1<-spplot(nc.province, c(paste("Incidence",i,sep="")),main=paste("Incidence ",q4[i],"-",(q4[i+1]-1)," days", sep=""),col.regions =cols[c1i:c1f]) 
i<-2
p2<-spplot(nc.province, c(paste("Incidence",i,sep="")),main=paste("Incidence ",q4[i],"-",(q4[i+1]-1)," days", sep=""),col.regions =cols[c2i:c2f])
i<-3
p3<-spplot(nc.province, c(paste("Incidence",i,sep="")),main=paste("Incidence ",q4[i],"-",(q4[i+1]-1)," days", sep=""),col.regions =cols[c3i:c3f])
i<-4
p4<-spplot(nc.province, c(paste("Incidence",i,sep="")),main=paste("Incidence ",q4[i],"-",(q4[i+1])," days", sep=""),col.regions =cols[c4i:c4f])

gc()
library(gridExtra)
grid.arrange(
p1,p2,p3,p4,nrow = 2)
```

##################predictions 3 days forward of new CODID 19 cases

```{r}
Forecast=3
dat_csv2<-dat_csv 
dat_csv2$t<-rep(1:days+days,nprov)
dat_csv2$totale_casi<-NA
dat_csv2<-dat_csv2[dat_csv2$t<=(days+Forecast),]
dat_csv_n<-rbind(dat_csv,dat_csv2)
dat_csv_n$t3<-dat_csv_n$t2<-dat_csv_n$t
#introducting hubei starting on 22/01, quarantine started on 25/01
# applying a smoothing , data has errors
hubei_new_cases<-filter(diff(hubei), rep(1, 5))

plot(hubei/58.5,type="l",ylab="COVID19 cases for milion of inhabitant",ylim=c(0,1530),xlab="Days since over 100 cases")
lines(c(rep(NA,2),cumsum(tapply(dat_csv_n$totale_casi,dat_csv_n$t,sum))/10),col=2)
abline(v=2,col=1,lty=2)
abline(v=4,col=2,lty=2)
abline(v=10,col=3,lty=3)
legend("bottomright",c("Hubei","Lombardy"),lty=1,col=1:2)
text(10,1500,"Hubei quarantine")
text(12,1350,"Lodi home lock",col=2)
text(20,1200,"Lombardy home lock",col=3)
#province of lodi restrictive measure started on 24/02 lag 28 days, at the beginning of the time series
#other provinces on 11/03 lagged started their restrictions
hubei_lodi<-hubei_new_cases[1:(days+Forecast)]
hubei_others<-c(rep(NA,8),hubei_new_cases[1:(days+Forecast-8)])
dat_csv_n<-dat_csv_n[order(dat_csv_n$codice_provincia,dat_csv_n$t),]
dat_csv_n$hubei<-rep(hubei_others,12)
dat_csv_n$hubei[dat_csv_n$denominazione_provincia=="Lodi"]<-hubei_lodi
#overdispersion
dat_csv_n$over<-1:dim(dat_csv_n)[1]
```

We modelled the number of cases.

In the rw2 I set constr = FALSE is set to FALSE and that, for this reason, the intercept is not included in the linear predictor.

```{r}
## include pop e hubei
formula.int_n<-  totale_casi ~ 1+log(hubei+1)+log(pop)+ f(ID,model="bym",graph=file.adj) +
f(t,model="rw2") + f(t2,model="iid") +
f(t3,model="iid", group=ID2, control.group=list(model="besag", graph=file.adj))

fit_n<-inla(formula.int_n, family="poisson", data=dat_csv_n,control.compute = list(dic=T),control.predictor = list(link = 1))
summary(fit_n)
```

```{r}
#### components
#overall trend
t_1<-fit_n$summary.random$t$mean 
#iid trend
t_2<-fit_n$summary.random$t2$mean 
 # spatial CAR componen
s_1<-fit_n$summary.random$ID$mean[1:nprov]
# spatial idd
s_2<-fit_n$summary.random$ID$mean[(nprov+1):(nprov*2)] 
# spatio-temporal coefficients 
st<-fit_n$summary.random$t3$mean
## Trends for province
Date_n<-seq(as.Date("2020-02-24"),as.Date("2020-02-24")+days+Forecast-1,1)
trends<-t_1+t(matrix(st,nrow=nprov))
plot(Date_n,t_1,ylab="RW2 Time coefficient",xlab="Date")
matplot(t(matrix(st,nrow=nprov)),ylab="RW2 Time coefficient for Province",xlab="Date")
#extract number of cases for lodi
est_lodi<-fit_n$summary.fitted.values[dat_csv_n$denominazione_provincia=="Lodi",]
